---
title: "HW1-OLS_Regression"
output: html_document
date: "2023-10-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(kableExtra)

options(scipen = 999)
```

# Introduction

# Methods
## Data Cleaning

## Exploratory Data Analysis

## Multiple Regression Analysis

## Additional Analyses

## Software

# Results
## Exploratory Results

```{r data}

data <- read.csv("Data/RegressionData.csv")

```

```{r histogram}

hist(data$MEDHVAL,breaks = 50)
hist(data$PCTBACHMOR, breaks = 50)
hist(data$NBELPOV100, breaks = 50)
hist(data$PCTVACANT, breaks = 50)
hist(data$PCTSINGLES, breaks = 50)

```

```{r table}

summary_stats_mean <- data %>%
  summarise(HEDVAL = mean(MEDHVAL),
            PCTBACHMOR = mean(PCTBACHMOR),
            NBELPOV100 = mean(NBELPOV100),
            PCTVACANT = mean(PCTVACANT),
            PCTSINGLE = mean(PCTSINGLES)) %>%
  gather(key = "variable", value = "mean")
            
summary_stats_sd <- data %>%
  summarise(HEDVAL = sd(MEDHVAL),
            PCTBACHMOR = sd(PCTBACHMOR),
            NBELPOV100 = sd(NBELPOV100),
            PCTVACANT = sd(PCTVACANT),
            PCTSINGLE = sd(PCTSINGLES)) %>%
  gather(key = "variable", value = "sd") %>%
  mutate(row_names = c('Median House Value','# Households Living in Poverty','% of Individuals with Bachelor Degrees or Higher','% of Vacant Houses','% of Single House Units')) 

summary_stats

left_join(summary_stats_mean, summary_stats_sd, by='variable') %>%
  dplyr::select('row_names','mean','sd') %>%
  kbl(col.names = c('Variable','Mean','Standard Deviation')) %>%
  kable_classic()
  
```

# quick code that cehcks which variables have 0 values for logarithmic transformation 
```{r 0_value_check}
zero_columns <- apply(data, 2, function(col) any(col == 0)) 

variables_with_zero_values <- names(zero_columns[zero_columns])

cat("Columns with 0 values:", paste(variables_with_zero_values, collapse = ", "))

```

```{r log_trans}
LNMEDHVAL <- log(data$MEDHVAL)
LNPCBACHMORE <- log(1 + data$PCTBACHMOR)
LNNBELPOV100 <- log(1 + data$NBELPOV100)
LNPCTVACANT <- log(1 + data$PCTVACANT)
LNPCTSINGLES <- log(1 + data$PCTSINGLES)

hist(LNMEDHVAL)
hist(LNPCBACHMORE)
hist(LNNBELPOV100)
hist(LNPCTVACANT)
hist(LNPCTSINGLES)

```
 

# Scatter Plots - Looking at linearity between dependent variable (LNMEDHVAL) and predictors (LNNBELPOV100,data$PCTBACHMOR, data$PCTVACANT, data$PCTSINGLES  )  
 
```{r scatter}
plot(LNNBELPOV100, LNMEDHVAL)
plot(data$PCTBACHMOR,LNMEDHVAL)
plot(data$PCTVACANT,LNMEDHVAL)
plot(data$PCTSINGLES, LNMEDHVAL)
```



# Pearson correlations

```{r pearson}
predictors <- as.matrix(cbind( data$PCTBACHMOR, data$PCTVACANT,data$PCTSINGLES,LNNBELPOV100 ))

cor(predictors)
```
library(car)
vif_result <- vif(predictors)

## Regression Results

## Regression Assumptions Checks

## Additional Models

# Discussion and Limitations